Regression Cube Analysis of Cohort Study Data

### 2015-03-31 draft.tex
minor changes
further revisions

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
Furthermore, the insights can be used to derive recommendations regarding a healthy lifestyle or to provide information about wide-spread diseases.
%%
During the standard workflow, physicians transform observations into hypotheses, which are depicted using epidemiological features and then are statistically analyzed.

@@

Medical image data is also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not focus on analyzing image data.
%%
More discussions of cohort study data types and characteristics can be found in \cite{Klemm2014VIS, Preim2014, Toennies2015}.
%%
%\paragraph{Image Data.}
% \com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
% %%
% Since the Rotterdam study, many modern cohort studies include medical image data.
% %%
% For ethical reasons, the imaging modalities usually not include ionizing radiation.
% %%
% The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
% %%
% These data are hard to analyze as they require segmentations to outline structures of interest.
% %%
% This process is prone to inter- and intra-observer variability when carried out manually.
% %%
% Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
% %%
% Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).
\\\\
%\paragraph{Confounding Features.}
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%Confounding features have to be considered in epidemiological analysis, as they influence both the exposure as well as the outcome of an analysis.
%%
%Features influencing the exposure as well as the outcome of an analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

We defined an Interactive Visual Analysis workflow for image-centric cohort study data, by extending the feature selection step (recall \ref{sec:Background}) with an iterative analysis loop incorporating group selection and visualization using expert input as well as clustering methods \cite{Klemm2014VIS}.
%%
Information visualizations of cohort study features were augmented with extracted image data, yielding linked views with both image and non-image information.
%%
% erst die ErklâÂ§rung, dann die Folge daraus =)
The mosaic plot presented in \cite{Klemm2014VIS} highlights the pairwise correlation between features without a target disease by depicting the \emph{Cram\'{e}r's V} contingency values.
%%
Its great popularity among our domain experts was the inspiration for this work.
%%

@@

The CFS algorithm will select the feature which has the largest explanatory power and discards the other dimensions.

@@

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/workflow}
 \caption{
 Different workflow types of the analysis using \emph{Regression Cubes} (based on~\cite{Keim}).
 %%
 % The analysis starts with a set of features given by the data set.
 %%
 [\textbf{1.AF}] The workflow starts by declaring a formula to specify a hypothesis, or to use a predefined formula for a hypothesis-free analysis.
 %%
 [\textbf{2.SI}] The \emph{Regression Cube} is then visualized.
 %%
 The user has the option to either adjust the formula or to derive details on demand on a specific regression.
 %%
 [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or a selection of a difference view.
 %%
 The latter is used to compare regression cubes.
 %%
 [\textbf{4.DD}] Details about features using the 2D mosaic plot representation yield insights and hypotheses about feature relations.
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
% â¹bergang ist recht holprig gewesen
Our \emph{Regression Cube} model is well-suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al.~\cite{Keim}:

@@

\textbf{Encoding via csv Files.}
%%
Encoding variable types in CSV files is not standardized.
%%
However, we need to ensure the correct variable type classification and have to enforce some basic standards.
%%
All categorical values have to be enclosed by quotation marks.
%%
Continuous variables are denoted as digits without enclosing quotation marks.
%%
Although this seems obvious, many cohort study data sets encode categorical features using ID values that are denoted in a data dictionary.
%%
Variables with only two manifestations are classified as dichotomous, leading to three possible data types: numerical, categorical and categorical/dichotomous.
%%
Missing values are denoted by using no character at all, a whitespace, or an empty quotation mark encapsulated string.

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written in \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{Regression Cubes}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
% The number of attached machines to the system
%    the  the  Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
% Using web technologies for prototyping allows for:
% \begin{itemize}
% 	\item easy exchange of software, without the need of installing anything other than a modern web browser,
% 	\item fast iterative adaption of method changes, without the need of installing updates,
% 	\item outsourcing computation-heavy tasks to server-clusters, which do not have to be in local proximity.
% \end{itemize}
% %%
%%
We rely on web-based technologies for our prototype.
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
\texttt{Angular.js} is forcing developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
% UI muss mal erklärt ewrden, tauchte schon weiter vorn im Paper auf
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of UI elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Open source volume rendering methods are available but do not satisfy our requirements.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

@@

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\\\texttt{github.com/paulklemm/regression-cube-r-package}}}$^{,}$\footnote{Front-End and Node.js Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{\\github.com/paulklemm/regression-cube-prototype}}}
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical back-end through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku homepage.
%%
%The front-end source as well as the \texttt{Nodejs} backend can be found here: .
%%
% \com{Use VAST'14 as Guideline}
% \com{Cloud computing}
% \com{Volume Rendering vs Particles}
% \com{Speed is merely a matter of available server machines due to parallelization process.}
% \com{Security enabled by hashing files.}
% \com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al.~\cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We employ the \emph{VDAR} technique for analyzing our system.
%%
\paragraph{Participants, Setup and Procedure.}
%%
%One major advantage of focusing on web-based research was the possibility to bridge the geographical gap towards the epidemiological domain experts.
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

@@

%%
We conducted a three with two participants. %, who also co-authored this publication.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set analysis
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.

The hepatic steatosis data set is analyzed by \emph{UN}, a data scientists responsible for prior analysis of the data.
%%

% OK so?
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used in \cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
% ok?  to indicate absence or presence of the disease.
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present the last study wave.

%%
Apart from the target feature, the data set contains 199 features, comprising sociodemographic features (gender, age), consumption behavior (e.g., alcohol or tobacco), laboratory data (e.g., sera concentrations), and two features depicting the liver ultrasound.
%%
%The acquisition wave is denoted using the appendix; 85 for S0 (SHIP-0), 55 for S1, 55 for S2 and 10 time-independent SNPs (DNA base pairs).
%%
The acquisition wave is denoted using the appendix; \emph{s0} meaning that their values have been recorded in SHIP-0 (first study moment), 50 features for \emph{s1} and 55 for \emph{s2}, alongside with 10 time-independent SNPs (DNA base pairs).
%%

In \cite{Niemann2014}, the authors show different class distributions of liver fat concentrations of women and men.
%%
For women, an association between age and liver fat was identified.
%Also, for women they identify that there is a association between age and liver fat concentration.
%%
An appropriate cut-off value of 52 years was set, yielding the most homogeneous class distribution within the resulting subsets.
%%
% ist das richtig? nicht female<52,male, female>52?
Based on these observations, we perform our analysis on three populations: \emph{males}, \emph{females} (all ages) \emph{females older than 52 years}.
%%
\subsection{The Breast Fat Data Set}
The breast fat data set was compiled to find associations between the parenchyma tissue proportion in the female breast compared to other features in the data.
%%
The ratio between parenchyma and cellular connective tissue (breast density) has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

The data comprises 1,186 female subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about somatometric features (e.g., body size or weight) consumption behavior (e.g., alcohol or tobacco), personal and medical history (e.g., occupation or prior diseases), women-specific features (e.g., number of born children or contraception type) as well as mammography features (e.g., fat content or parenchyma tissue proportion to volume).
% It contains 231 dimensions, holding information about
% \begin{itemize}
% 	\item \emph{somatometric features}, e.g., body size, weight or BMI,
% 	\item \emph{lifestyle features}, e.g., alcohol/tobacco consume,
% 	\item \emph{personal history}, e.g., occupation, marital status,
% 	\item \emph{medical history}, e.g., current or prior diseases or laboratory values, such as cholesterol levels,
% 	\item \emph{women specific features}, e.g., number of born children, contraception type or hormone replacement therapy and
% 	\item \emph{mammography features} derived from the image data.
% \end{itemize}
%%
The latter were derived from manually segmented MRI data for each subject carried out by radiologists.
%%
%%
%\com{Describe Parenchym tissue and its relation to breast cancer.}
%\com{Describe how the data was acquired and preprocessed}

@@

Analyzing the mosaic plot [\textbf{4.DD}] revealed high correlations for somatometric features, hepatic steatosis indicator features as well as laboratory values, such as \emph{creatinine} (used renal retention parameter) and \emph{uric acid} (among others used as gout and diabetes risk factors) magnitudes.

@@

\textbf{Interleukin-6 Correlation With Liver Fat.}
%%
During the analysis, one hotspot was always observable in the [\textbf{2.SI}] and [\textbf{3.ZF}] incorporating a high \emph{Interleukin-6} (\emph{IL-6}) correlation with liver fat values ($R^2$ of $0.8$, see Fig.~\ref{fig:Application}b).
%%
The correlation was high for both the dichotomized and continuous target feature.
%%
The literature described relations between \emph{IL-6} and liver cancer \cite{He2013} as well as chronic liver diseases \cite{Streetz2003}.
%%
For mice, strong effects of \emph{IL-6} with hepatic steatosis were described \cite{Hong2004}.
%%
The finding is subject of further analysis.
%%
\subsection{Case 2: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis aims to find relationships on the breast fat data using mammography analysis features.
%%
Relationships between the share parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%
The [\textbf{1.AF}] was started by \emph{KH} using the default formula for hypothesis-free analysis ($Z \sim X + Y$).
%%
At first, she was interested in correlations with the \emph{parenchyma tissue} percentage, which was selected through the drop-down for the z-axis [\textbf{2.SI}].
%%
She observed strong correlations with \emph{age}, \emph{body fat percentage}, \emph{hip} and \emph{waist circumference} as well as \emph{menstrual period} or \emph{pregnancy status} as expected (Fig.~\ref{fig:Application} right).
%%
Women with higher \emph{body fat} also have a larger \emph{breast fat percentage}, which also correlates with other somatometric features.
%%
\emph{Age} is a strong influencing factor, as breast tissue and subsequently the parenchyma tissue degrades over time.
%%
\emph{KH} proceeded using [\textbf{3.ZF}] and [\textbf{4.DD}] to check for relationships for different target features, such as current \emph{hormone replacement therapy}, \emph{BI-RADS} (classification of the mammography findings) as well as different diseases, such as \emph{diabetes} or \emph{gout}.
%%
She observed relationships matching her expectations and expert knowledge.
%%
One unexpected relationship was observed between \emph{breast lesions} and \emph{menstruation cycle} towards \emph{spiral contraception} ($R^2$ of $0.77$).
%%
\emph{KH} proceeded with a detailed analysis of the parenchyma tissue.
%\emph{KH} derived at the hypotheses: \emph{Somatometric features impose the major influences on the parenchyma tissue.}
%%

>>>>

\maketitle
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
Furthermore, the insights can be used to derive recommendations regarding a healthy lifestyle or to provide information about widespread diseases.
%%
During the standard workflow, physicians transform observations into hypotheses, which are depicted using epidemiological features and then are statistically analyzed.

@@

Medical image data is also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not focus on analyzing image data.
%%
More discussions of cohort study data types and characteristics can be found in \cite{Preim2014, Toennies2015}.
%%
%\paragraph{Image Data.}
% \com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
% %%
% Since the Rotterdam study, many modern cohort studies include medical image data.
% %%
% For ethical reasons, the imaging modalities usually not include ionizing radiation.
% %%
% The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
% %%
% These data are hard to analyze as they require segmentations to outline structures of interest.
% %%
% This process is prone to inter- and intra-observer variability when carried out manually.
% %%
% Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
% %%
% Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).
\\\\
%\paragraph{Confounding Features.}
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%Confounding features have to be considered in epidemiological analysis, as they influence both the exposure as well as the outcome of an analysis.
%%
%Features influencing the exposure as well as the outcome of an analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

We defined an Interactive Visual Analysis workflow for image-centric cohort study data, by extending the feature selection step (recall \ref{sec:Background}) with an iterative analysis loop incorporating group selection and visualization using expert input as well as clustering methods \cite{Klemm2014VIS}.
%%
Information visualizations of cohort study features were augmented with extracted image data, yielding linked views with both image and non-image information.
%%
The mosaic plot presented in \cite{Klemm2014VIS} highlights the pairwise correlation between features without a target disease by depicting the \emph{Cram\'{e}r's V} contingency values.
%%
Its great popularity among our domain experts was the inspiration for this work.
%%

@@

The CFS algorithm will select the feature, which has the largest explanatory power and discards the other dimensions.

@@

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/workflow}
 \caption{
 Different workflow types of the analysis using \emph{Regression Cubes} (based on~\cite{Keim}).
 %%
 % The analysis starts with a set of features given by the data set.
 %%
 [\textbf{1.AF}] The workflow starts by declaring a formula to specify a hypothesis, or to use a predefined formula for a hypothesis-free analysis.
 %%
 [\textbf{2.SI}] The \emph{Regression Cube} is then visualized.
 %%
 The user has the option to either adjust the formula or to derive details on demand on a specific regression.
 %%
 [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or a selection of a difference view.
 %%
 The latter is used to compare regression cubes.
 %%
 [\textbf{4.DD}] Details about features using the 2D mosaic plot representation yield insights and hypotheses about feature relations.
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
% â¹bergang ist recht holprig gewesen
Our \emph{Regression Cube} model is well suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al.~\cite{Keim}:

@@

\textbf{Encoding via CSV Files.}
%%
Encoding variable types in CSV files is not standardized.
%%
However, we need to ensure the correct variable type classification and have to enforce some basic standards.
%%
All categorical values have to be enclosed by quotation marks.
%%
Continuous variables are denoted as digits without enclosing quotation marks.
%%
Although this seems obvious, many cohort study data sets encode categorical features using ID values that are denoted in a data dictionary.
%%
Variables with only two manifestations are classified as dichotomous, leading to three possible data types: numerical, categorical and categorical/dichotomous.
%%
Missing values are denoted by using no character at all, a whitespace, or an empty quotation mark encapsulated string.

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written using \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{Regression Cubes}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
% The number of attached machines to the system
%    the  the  Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
% Using web technologies for prototyping allows for:
% \begin{itemize}
% 	\item easy exchange of software, without the need of installing anything other than a modern web browser,
% 	\item fast iterative adaption of method changes, without the need of installing updates,
% 	\item outsourcing computation-heavy tasks to server-clusters, which do not have to be in local proximity.
% \end{itemize}
% %%
%%
We rely on web-based technologies for our prototype.
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
It forces developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
% UI muss mal erklärt ewrden, tauchte schon weiter vorn im Paper auf
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of UI elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Open source volume rendering methods are available but do not satisfy our requirements.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

@@

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\\\texttt{github.com/paulklemm/regression-cube-r-package}}}$^{,}$\footnote{Front-End and Node.js Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{\\github.com/paulklemm/regression-cube-prototype}}}
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical back-end through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku app.
%%
%The front-end source as well as the \texttt{Nodejs} backend can be found here: .
%%
% \com{Use VAST'14 as Guideline}
% \com{Cloud computing}
% \com{Volume Rendering vs Particles}
% \com{Speed is merely a matter of available server machines due to parallelization process.}
% \com{Security enabled by hashing files.}
% \com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al.~\cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, Setup and Procedure.}
%%
%One major advantage of focusing on web-based research was the possibility to bridge the geographical gap towards the epidemiological domain experts.
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

@@

%%
We conducted the analysis with three participants. %, who also co-authored this publication.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set analysis.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientists responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used in \cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present the last study wave.

%%
Apart from the target feature, the data set contains 199 features, comprising sociodemographic features (gender, age), consumption behavior (e.g., alcohol or tobacco), laboratory data (e.g., sera concentrations), and two features depicting the liver ultrasound.
%%
%The acquisition wave is denoted using the appendix; 85 for S0 (SHIP-0), 55 for S1, 55 for S2 and 10 time-independent SNPs (DNA base pairs).
%%
The acquisition wave is denoted using the appendix; 85 features with appendix \emph{s0} denote their affiliation to SHIP-0 (first study moment), 50 features for \emph{s1} and 55 for \emph{s2}, alongside with 10 time-independent SNPs (DNA base pairs).
%%

In \cite{Niemann2014}, the authors show different class distributions of liver fat concentrations of women and men.
%%
For women, an association between age and liver fat was identified.
%Also, for women they identify that there is a association between age and liver fat concentration.
%%
An appropriate cut-off value of 52 years was set, yielding the most homogeneous class distribution within the resulting subsets.
%%
% ist das richtig? nicht female<52,male, female>52?
Based on these observations, we perform our analysis on three populations: \emph{males}, \emph{females (all ages)} and \emph{females older than 52 years}.
%%
\subsection{The Breast Fat Data Set}
The breast fat data set was compiled to find associations between the parenchyma tissue proportion in the female breast compared to other features in the data.
%%
The ratio between parenchyma and cellular connective tissue (breast density) has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

The data comprises 1,186 female subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about somatometric features (e.g., body size or weight) consumption behavior (e.g., alcohol or tobacco), personal and medical history (e.g., occupation or prior diseases), women-specific features (e.g., number of born children or contraception type) as well as mammography features (e.g., fat content or parenchyma tissue proportion to volume).
% It contains 231 dimensions, holding information about
% \begin{itemize}
% 	\item \emph{somatometric features}, e.g., body size, weight or BMI,
% 	\item \emph{lifestyle features}, e.g., alcohol/tobacco consume,
% 	\item \emph{personal history}, e.g., occupation, marital status,
% 	\item \emph{medical history}, e.g., current or prior diseases or laboratory values, such as cholesterol levels,
% 	\item \emph{women specific features}, e.g., number of born children, contraception type or hormone replacement therapy and
% 	\item \emph{mammography features} derived from the image data.
% \end{itemize}
%%
The latter were derived from MRI data for each subject, which was manually segmented by radiologists.
%%
%%
%\com{Describe Parenchym tissue and its relation to breast cancer.}
%\com{Describe how the data was acquired and preprocessed}

@@

Analyzing the mosaic plot [\textbf{4.DD}] revealed high correlations for somatometric features, hepatic steatosis indicator features as well as laboratory values, such as \emph{creatinine} (used as renal retention parameter) and \emph{uric acid} (among others used as gout and diabetes risk factors) magnitudes.

@@

\textbf{Interleukin-6 Correlation With Liver Fat.}
%%
During the analysis, one hotspot was always observable in the [\textbf{2.SI}] and [\textbf{3.ZF}] steps, incorporating a high \emph{Interleukin-6} (\emph{IL-6}) correlation with liver fat values ($R^2$ of $0.8$, see Fig.~\ref{fig:Application}b).
%%
The correlation was high for both the dichotomized and continuous target feature.
%%
The literature described relations between \emph{IL-6} and liver cancer \cite{He2013} as well as chronic liver diseases \cite{Streetz2003}.
%%
For mice, strong effects of \emph{IL-6} with hepatic steatosis were described \cite{Hong2004}.
%%
The finding is subject of further analysis.
%%
\subsection{Case 2: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis aims to find relationships on the breast fat data using mammography analysis features.
%%
Relationships between the share parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%
The [\textbf{1.AF}] was started by \emph{KH} using the default formula for hypothesis-free analysis ($Z \sim X + Y$).
%%
At first, she was interested in correlations with the \emph{parenchyma tissue} percentage, which was selected through the drop-down for the z-axis [\textbf{2.SI}].
%%
She observed strong correlations with \emph{age}, \emph{body fat percentage}, \emph{hip} and \emph{waist circumference} as well as \emph{menstrual period} or \emph{pregnancy status} as expected (Fig.~\ref{fig:Application} right).
%%
Women with higher \emph{body fat} also have a larger \emph{breast fat percentage}, which also correlates with other somatometric features.
%%
\emph{Age} is a strong influencing factor, as breast tissue and subsequently the parenchyma tissue degrades over time.
%%
\emph{KH} proceeded using [\textbf{3.ZF}] and [\textbf{4.DD}] to check for relationships for different target features, such as current \emph{hormone replacement therapy}, \emph{BI-RADS} (classification of the mammography findings) as well as different diseases, such as \emph{diabetes} or \emph{gout}.
%%
She observed relationships matching her expectations and expert knowledge.
%%
One unexpected relationship was observed between \emph{breast lesions} and \emph{menstruation cycle} towards \emph{spiral contraception} ($R^2$ of $0.77$).
%%
\emph{KH} proceeded with a detailed analysis of the parenchyma tissue.
%\emph{KH} derived at the hypotheses: \emph{Somatometric features impose the major influences on the parenchyma tissue.}
%%
